Microlensing as a possible probe of event-horizon structure in quasars

### 2016-07-18 paper1/ms.tex
editing ... unstable

\subsection{Rotationally symmetric source with a bivariate gaussian surface brightness distribution}

@@

Having used the simple model of an ideal fold to gain insight, we now
move to the numerical computation of microlensing light curves with
realistic mass distributions, and then fit source models to
lightcurves in the presence of random and systematic errors.

First we generate microlensing magnification maps using the
microlensing code created by Joachim Wambsgans (\citep{1999A&A...346L...5W, 1990PhDT.......180W, 1999JCoAM.109..353W}, 
\cite{1990PhDT.......180W}, \cite{1999JCoAM.109..353W}). We then generate
mock light curves for three source models --- a crescent, a uniform
disc and a Gaussian disc --- and then fit each mock light curve to all
three source models. Parameter fitting and marginalising was done by
Markov-chain Monte Carlo (MCMC).

@@

\subsection{Numerical microlensing magnification maps}

The code uses a ray-shooting technique to compute the gravitational
lensing effect of a mass distribution consisting of (a)~a smooth
component and (b)~a random distribution of point masses representing
stars.  The ray shooting maps a grid of $\vec\theta$ to $\vec\beta$
using the lens equation (\ref{eqn:lens}).  That is, the rays are shot
from the observer back to the source.

For the computation of the individual deflection angles
$\vec\alpha(\vec\theta)$ a hierarchical or tree method is used.  The
positions of all lensing masses are put into a grid of $\vec\theta$.
Each grid cell is subdivided into four smaller squares recursively
until every cell contains only one mass.  Nearby masses are added
individually while distance masses are clumped into larger grid cells
whose net contribution is approximated by its first few multipole
moments.  Scaled units are used, with the constant pre-factor in the
deflection angle (\ref{eqn:alpha}) separated out.

The result of ray shooting is a pixel map on the $\vec\beta$ plane of
the number of lightrays which arrive at the source plane from a
particular observer.  This intermediate result is effectively a
magnification map on the source plane.  Once the map is created, the
lightcurve can be obtained by specifying the transit path of the
source across the map.  At each point on this transit line, the code
computes equation (\ref{eqn:ft2d}) convolving the brightness
distribution of the source with the magnification pattern of the map.
In real life, not only are both lens and source moving but the lens
lens configuration, and with it the magnification pattern, is also
changing with time.  While the first subtlety is taken care of by a
coordinate transformation in this analysis, for the second one the
lens configuration is assumed to be constant in time.

When generating the magnification map depicted in
Figure~\ref{fig:magnification_map}, which is used in our analysis
below, only two point masses were included.  This was done in order to
have clean fold caustics.  For the computation of the actual
lightcurve, the code was modified to also allow for crescent shaped
images specified through the parameter set $R_p,R_n,a,b$. Here $R_p$
denotes the outer radius of the crescent and $R_n$ the inner one. The
orientation of the source image with respect to the magnification map
is specified by the parameters $a$ and $b$ as the shift of the center
of the inner disk from the center of the outer disk in $x$- and
$y$-direction respectively. In the original version of the code,
gaussian and disk-shaped images were already implemented. Those are
completely characterised by the single parameter $R_p$. The values of
the parameters are specified in pixel units corresponding to the
magnification map. Further, one needs to specify the start and end
point coordinates of the path, which the center of the source image
follows through the magnification map (see the depiction in figure
\ref{fig:magnification_map}). Hence, the points along this path are
specified through the number of time steps for which the computation of
the brightness is to be carried out. Those points correspond to the
actual measurement of the brightness of an object in the observational
case. For each timestep the two-dimensional convolution of equation
\ref{eqn:ft2d} is carried out numerically for the position of the
source on the magnification map.

For the purpose of this analysis, it was desirable to mimic the
analytical behaviour of a simple fold as much as possible, for
comparing the numerical result with the analytical one, therefore, the
path of the source was chosen, so that it intersects the border of the
caustic perpendicularly, and on a point where the border is a fold
caustic.

\begin{figure}
\centering
\includegraphics[width=0.9\hsize]{figures/data_cc.eps}
\includegraphics[width=0.9\hsize]{figures/data_dd.eps}
\includegraphics[width=0.9\hsize]{figures/data_gg.eps}
\caption{\label{fig:mockdata} Three different mock datasets with
  three different parametrizations fitted to each --- crescent (red),
  uniform-disk (blue) and Gaussian disk (green).}
\end{figure}

\subsection{Mock light curves}

Mock light curves were generated by running a source across a
caustic, specifically along the line from point C to point B in
Figure~\ref{fig:magnification_map}.  The model light curves, to which
these are fitted, are made by running a source across a different
caustic, along the line from point A to point B in the same figure.
That is, the mock data are generated with a clean but not ideal fold
caustic and then fitted with another such caustic.  This mimics the
unavoidable systematic error of not knowing the caustic exactly.

The crescent source had parameters (as explained in
\S\ref{subsec:crescent}) $R_p$ and $R_n$ being the radii of the outer
and inner circles and $(\alpha,\beta)$ being the coordinates of the
inner circle with respect to the centre of the outer circle in the
coordinate system associated with the image in Figure 1 and not
associated with the caustic surface as defined in the previous
sections.  The parameter values were
\begin{equation}
   (R_p, R_n, \alpha, \beta) = (50.0, 30.0, 15.0, 10.0).
\label{eqn:cp}
\end{equation}
The uniform disc had the same (outer) radius as the crescent.  In the
Gaussian source, we set $3\sigma=50$.  For convenience, below we will
refer to $R_p$ of the Gaussian source, by which we actually mean
$3\sigma$.

Figure~\ref{fig:mockdata} shows the three light curves.  Each has 250
points regularly spaced in time, with Gaussian noise at the level of
10\% of the current brightness.  This amounts formally to a single
summary data point with a signal to noise of
$10\times\sqrt{250}\simeq160$.

\subsection{Model fitting and likelihood analysis}

>>>>

\subsection{Rotationally symmetric source with a bivariate gaussian surface brightness distribution}\label{subsec:gaussian}

@@

Having used the simple model of an ideal fold to gain insight, we now
consider the question of whether the characteristics of a crescent
source could be discerned in real data having noise and systematic
error as well.  Our method is to generate microlensing lightcurves
\citep[using the code from][with a small extension to provide the
  option of crescent sources]
      {1990PhDT.......180W,1999A&A...346L...5W,1999JCoAM.109..353W}
      and then fitting these mock lightcurves to source models using
      Markov-chain Monte Carlo (MCMC).

\subsection{Numerical microlensing lightcurves}

The microlensing code uses a ray-shooting technique to compute the
gravitational lensing effect of a mass distribution consisting of
(a)~a smooth component and (b)~a random distribution of point masses
representing stars.  The ray shooting maps a grid of $\vec\theta$ to
$\vec\beta$ using the lens equation (\ref{eqn:lens}).  That is, the
rays are shot from the observer back to the source.  For the
computation of the individual deflection angles
$\vec\alpha(\vec\theta)$ a hierarchical or tree method is used.  The
positions of all lensing masses are put into a grid of $\vec\theta$.
Each grid cell is subdivided into four smaller squares recursively
until every cell contains only one mass.  Nearby masses are added
individually while distance masses are clumped into larger grid cells
whose net contribution is approximated by its first few multipole
moments.  Scaled units are used, with the constant pre-factor in the
deflection angle (\ref{eqn:alpha}) separated out.  The result of ray
shooting is a pixel map on the $\vec\beta$ plane of the number of
lightrays which arrive at the source plane from a particular observer.
This intermediate result is effectively a magnification map on the
source plane.

When generating the magnification map depicted in
Figure~\ref{fig:magnification_map}, which is used in our analysis
below, only two point masses were included.  This was done in order to
have clean fold caustics.  There was no smooth component.

Once the map is created, the lightcurve can be obtained by specifying
the transit path of the source across the map.  The source is moved
along the transit line, one pixel at a time, and at each step the
brightness distribution of the source is convolved
(equation~\ref{eqn:ft2d}) with the magnification map to give the
brightness.  In real life, not only are both lens and source moving
but the lens lens configuration, and with it the magnification
pattern, is also changing with time.  While the first subtlety is
taken care of by a coordinate transformation in this analysis, for the
second one the lens configuration is assumed to be constant in time.

For the purpose of this analysis, it was desirable to mimic the
analytical behaviour of a simple fold as much as possible, for
comparing the numerical result with the analytical one, therefore, the
path of the source was chosen, so that it intersects the border of the
caustic perpendicularly, and on a point where the border is a fold
caustic.  We choose the line AC in Figure~\ref{fig:magnification_map}
and consider its two caustic-crossing segments separately:
\begin{itemize}
\item ``path~I'' runs from C to B;
\item ``path II'' runs from A to B.
\end{itemize}
The magnification along paths I and II are very similar, but not
identical.

We considered two sources, as follows.
\begin{itemize}
\item The crescent source (as explained in \S\ref{subsec:crescent})
  with parameters
\begin{equation}
   (R_p, R_n/Rp, a/Rp, b/Rp) = (1.31, 0.4, 0, 0.3)
\label{eqn:cp}
\end{equation}
corresponding to $r_{1/2}=1$.
\item A Gaussian source (as explained in \S\ref{subsec:crescent}),
  also with $r_{1/2}=1$.
\end{itemize}
Unit length has 30 pixels on the magnification map and 30 points on a
light curve.

Each lightcurve has 251 points regularly spaced in time, with Gaussian
noise at the level of 4\% of the current brightness.

Mock data were generated by taking path-I lightcurves for

This mimics the unavoidable systematic error of not knowing the
caustic exactly.

@@

\subsection{Model-fitting results}
